% !TEX root = ./main.tex
\section{Implementing the evolutionary forces}

As hinted at in Fig.~\ref{fig01:moran}(A), the Moran process gives us a simple
recipe for how to encode the different evolutionary forces on the transition
rates $W^\pm(x)$. Recall that since the transition rates in allele frequency
space are the same as the transition rates in number of organisms space, we can
more easily conceptualize the forces in the latter. In other words, the rate to
transition from $x$ to $x + \Delta x$ is by construction equal to the rate to
transition from $n$ to $n + 1$; therefore we can define the transition rates in
the more convenient space of organism number. In the Moran process changes in
population composition happen when one of the organisms in the population chosen
randomly dies and it is immediately replaced by an organism with a different
allele (since replacement with the same allele does not change the population
composition). This means that the general form in which the number of organisms
with allele $A$ can change looks like
\begin{equation}
    W^+(x) = w^+(n) = 
    \text{Rate of $A$ dying} \times
    \text{Probability of replacement by $a$},
\end{equation}
and 
\begin{equation}
    W^-(x) = w^-(n) = 
    \text{Rate of $a$ dying} \times
    \text{Probability of replacement by $A$},
\end{equation}
Having these general forms let us define the rates for different evolutionary
forces.

\subsection{Genetic drift}
\mrm{Further discussion of what genetic drift is?}

Given the intrinsic stochasticity of the Moran process, the easiest evolutionary
force to implement is that of genetic drift. For simplicity, we assume that both
allele types have the same death rate per organism, $\gamma$. For the case where
only genetic drift is changing the population composition, we assume that the
reproduction probability per organism is the same (equivalent to saying both
genotypes have the same fitness), so the probability of an organism that died
being replaced by an organism with the opposite allele is given by the relative
frequency of such allele. Mathematically this means that we can express the rate
with which the number of organisms with allele $A$ increases as
\begin{equation}
    W^+(x) = w^+(n) = 
    \overbrace{\gamma \times (N-n)}^{\text{rate of $a$ dying}} \times
    \overbrace{\frac{n}{N}}^
    {\substack{\text{prob. of $A$} \\ \text{replacing}}},
\end{equation}
where the term $\gamma \times (N-n)$ defines the rate at which an organism with
an allele $a$ dies, and the term $n/N$ defines the probability of an organism
with allele $A$ reproducing. Equivalently the rate with which the number of
organisms with allele $a$ decreases takes the form
\begin{equation}
    W^+(x) = w^+(n) = 
    \overbrace{\gamma n}^{\substack{\text{rate of $A$}\\ \text{dying}}}
    \overbrace{\frac{(N-n)}{N}}^
    {\substack{\text{prob. of $a$} \\ \text{replacing}}}.
\end{equation}
Given that both rates are equal, we can see that the first term in
Eq.~\ref{eq:pde_x_general} involving $W^+(x) - W^-(x)$ is zero. This means that
there are no deterministic (directional) forces when only genetic drift is
considered, just pure randomness. Substituting the sum of the rates into the
second term of Eq.~\ref{eq:pde_x_general} results in
\begin{equation}
\begin{aligned}
    \frac{\partial P(x, t)}{\partial t} &=\frac{1}{2 N^{2}} 
    \frac{\partial^{2}}{\partial x^{2}}[2 \gamma(N-n) n P(x, t)], \\
    &=\frac{\partial^{2}}{\partial x^{2}}
    \left[\frac{1}{2 N^{2}} 2 \gamma\left(\frac{(N-n) n}{N} \right) P(x, t)\right],\\
    &=\frac{\gamma}{N} \frac{\partial^{2}}{\partial x^{2}}[x(1-x) P(x, t)],
\end{aligned}
\end{equation}
where for the last step we substitute the definition of $x\equiv n/N$. We
redefine the time scale to be in units of $\gamma^{-1}$, meaning that the time
units are measured in terms of the mean life expectancy of an organism. This
allows us to write
\begin{equation}
    \frac{\partial P(x, t)}{\partial t} =
    \frac{1}{N} \frac{\partial^{2}}{\partial x^{2}}[x(1-x) P(x, t)],
\end{equation}
the classic Kimura diffusion equation for genetic drift only.

\subsection{Genetic drift plus selection}

Natural selection is intrinsically associated with the concept of fitness. The
phrase ``survival of the fittest'' first used by Darwin guided and still guides
the way that biologists think about the evolution of many organisms. But despite
the fact that fitness is part of the daily jargon of many biologists, it is a
subtle and highly debated concept. After all, what defines the ability of an
organism to survive the challenges that surround them are completely context
dependent. Roughly speaking we can think of fitness as the ability of an
organism, or a population of organisms to survive and reproduce in the given
ecological niche they occupy. The concept of fitness only makes sense when there
is a competition between organisms. The intrinsic growth rate of a completely
homogeneous population is irrelevant since, as we mentioned before, a population
with a single allele growing is not evolving since there are no changes to the
population composition. The term ecology has to be included because fitness is a
result of the interplay between organisms with their environment including all
biotic and abiotic interactions. 

It is common both in theory and in experiments to use the relative growth rates
of organisms, i.e. the speed at which they can reproduce and generate
offspring, as a proxy for fitness. This is a convenient approximation both
for experiments and for theory, but one should not lose track of the relevant
context dependence on the fitness. Just because redwoods have an average life
span of 500-700 years and a very low growth rate that doesn't mean they are not
fit. Having said that we will first begin with the simplest form of fitness,
i.e. frequency independent selection. The term frequency independence simply
refers to the assumption that the fitness of a particular allele does not
depend on the relative abundance of such allele. This assumption could break
down for cases such as some pathogenic bacteria that coordinate their attack
via cell-to-cell communication known as quorum sensing. 

To implement the effect of different reproductive success for different alleles
we introduce parameters $f_A$ and $f_a$ as the fitness values for allele $A$ and
$a$, respectively. With these parameters in hand we must redefine the
probability of an organism reproducing to replace the one that dies in the Moran
process. The replacement probability for allele $A$ is now given by
\begin{equation}
    \text{Prob. of $A$ replacing} = \frac{f_A n}{f_A n + f_a (N - n)}.
\end{equation}
Likewise for allele $a$ we have
\begin{equation}
    \text{Prob. of $a$ replacing} = \frac{f_a (N - n)}{f_A n + f_a (N - n)}.
\end{equation}
Let us now assume that $f_A \approx (1 + s) f_a$ for a small $s$. This parameter
$s$ is the so-called selection coefficient, which in nature can be of the order
of $10^{-3}$ or less. With these assumptions we can simplify these substitution
probabilities to be
\begin{equation}
    \text{Prob. of $A$ replacing} \approx \frac{n}{N}(1 + s),
\end{equation}
and
\begin{equation}
    \text{Prob. of $a$ replacing} \approx \frac{(N - n)}{N},
\end{equation}
where, after canceling $f_a$ from the numerator and denominator, we assume that
$(1 + s)n + (N - n) \approx N$ since $s \ll 1$. With these updated replacement
probabilities we again compute the population change rates $W^\pm(x)$ as the
product of the rate of certain type of organism dying times the probability of
being replaced by the opposite allele. These rates take the form
\begin{equation}
    W^+(x) = 
    \overbrace{\gamma \times (N - n)}^{\text{rate of $a$ dying}} \times
    \overbrace{\frac{n}{N} (1 + s)}^
    {\substack{\text{prob. of $A$ replacing}\\ \text{with fitness difference}}},
\end{equation}
and
\begin{equation}
    W^-(x) = \gamma n \frac{(N - n)}{N},
\end{equation}
With these rates we ca now compute the sum and difference required by
Eq.~\ref{eq:pde_x_general}. The difference of these two rates takes the form
\begin{equation}
    W^+(x) - W^-(x) = \gamma s \frac{n(N-n)}{N}.
\end{equation}
The sum results in
\begin{equation}
    W^+(x) + W^-(x) = \gamma (2 + s) \frac{n(N-n)}{N}.
\end{equation}
Substituting this into Eq.~\ref{eq:pde_x_general} results in
\begin{equation}
    \frac{\partial}{\partial t} P(x, t) =
    -\frac{1}{N} 
    \frac{\partial}{\partial x}
    \left[\gamma s \left(\frac{(N-n) n}{N}\right) P(x, t)\right] 
    +\frac{1}{2 N^{2}} 
    \frac{\partial^{2}}{\partial x^{2}}
    \left[\gamma (2+s)\left(\frac{(N-n) n}{N}\right) P(x, t)\right].
\end{equation}
Simplifying terms and substituting the definition of the allele frequency gives
\begin{equation}
    \frac{\partial}{\partial t} P(x, t) =
    -\gamma_{S} \frac{\partial}{\partial x}[x(1-x) P(x, t)] 
    +\frac{\gamma\left(1+\frac{s}{2}\right)}{N}
    \frac{\partial^{2}}{\partial x^{2}}[x(1-x) P(x, t)].
\end{equation}
To obtain the final form we again write the time scale in units of
$\gamma^{-1}$. Furthermore we use the simplification that $s \ll 1$, obtaining
the classic Kimura diffusion equation for selection and drift
\begin{equation}
    \frac{\partial}{\partial t} P(x, t) =
    -\frac{\partial}{\partial x}[s x(1-x) P(x, t)] 
    +\frac{1}{N} \frac{\partial^{2}}{\partial x^{2}}[x(1-x) P(x, t)].
\end{equation}

\subsection{Genetic drift plus selection plus mutation}

One of the ingredients for evolution to take place is the constant appearance of
genetic variability. After all, the raw material for evolution to act on is the
appearance of new mutations in the population. While both genetic drift and
selection reduce population diversity, mutation creates more diversity. In the
case of our one-locus two-allele case it can even resurrect alleles that wen
extinct as one organism changes its genetic content. The implementation of this
third force changes the possibilities on how to change the composition of the
population. As depicted in Fig.~\ref{fig01:moran}(A), if mutation is taken into
account, there are two possible substitutions which would modify the allele
frequency: (1) The usual path in which an organism with the opposite allele to
the one that died reproduces and does not mutate when doing so, and (2) the
possibility of an organism of the same allele as the one that died reproducing,
but when doing so, it mutates to the opposite allele. For simplicity we will
assume that the mutation probability from $A$ to $a$, $\mu_{A\rightarrow a}$, is
the same as from $a$ to $A$, $\mu_{a\rightarrow A}$. For simplicity let us
define the transition rate $W^+(x)$ as
\begin{equation}
    W^+(x) = W^+_{(1)}(x) + W^+_{(2)}(x),
\end{equation}
where we break the rate into the two possible paths. The first path in which an
organism of the opposite allele to the one that died replaces it and does not
mutate when doing so takes the form
\begin{equation}
    W^{+}_{(1)}(x) = 
    \overbrace{\gamma(N-n)}^{\text{rate of $a$ dying}}\times
    \overbrace{\frac{n}{N}(1+s)}^
    {\substack{\text{prob. of $A$ replacing}\\ \text{with fitness diff.}}}\times
    \overbrace{(1-\mu)}^
    {\substack{\text{prob. of not} \\ \text{mutating}}},
\end{equation}
where the evolutionary forces appear as a product of the rates and probabilities
of each of the steps taking place. For the second path in which the organism 
that replaces the one that dies is of the same type, but when it reproduces 
there is a mutation to the opposite allele, we have a rate of the form
\begin{equation}
    W^{+}_{(2)}(x) = 
    \overbrace{\gamma(N-n)}^{\text{rate of $a$ dying}}\times
    \overbrace{\frac{(N - n)}{N}}^
    {\text{prob. of $a$ replacing}}\times
    \overbrace{\mu}^{\text{prob. of mutating}}.
\end{equation}
Putting these two rates together results in a transition rate $W^{+}(x)$ of the
form
\begin{equation}
    W^{+}(x) = 
    \overbrace{\gamma(N-n) \frac{n}{N}(1+s) (1-\mu)}^{\text{path (1)}}
    + \overbrace{\gamma(N-n) \frac{(N-n)}{N} \mu}^{\text{path (2)}}.
\end{equation}
Equivalently, we can write the rate $W^{-}(x)$ as a decomposition of the two 
possible paths that lead to population structure changes. Putting these two 
rates back together results in
\begin{equation}
    W^{-}(x)=
    \overbrace{\gamma n \frac{(N-n)}{N}(1-\mu)}^{\text{path (1)}}
    +\overbrace{\gamma n \frac{n}{N}(1+s) \mu}^{\text{path (2)}}.
\end{equation}
Again we follow Eq.~\ref{eq:pde_x_general} and compute the sum and the
difference between these rates. After some algebra, we find that the difference
between these rates is of the form
\begin{equation}
    W^+(x) - W^-(x) = 
    \frac{\gamma}{N}\left[n s\left(N-n-N\mu\right)+N{\mu}(N-2 n)\right].
\end{equation}
For the sum of the rates we find
\begin{equation}
    W^+(x) + W^-(x) =
    \frac{\gamma}{N}\left[N n(2-4 \mu+s-\mu s) + 
    n^{2}(-2+4 \mu-s+2 \mu s)\right].
\end{equation}
Substituting these rates in Eq.~\ref{eq:pde_x_general} gives
\begin{equation}
\begin{split}
    \frac{\partial}{\partial t} P(x, t)=
    &-\frac{1}{N} \frac{\partial}{\partial x}
    \left[ \frac{\gamma}{N} \left(n s \left(N - n - N \mu \right)+
    N \mu (N - 2 n) \right) P(x, t)\right] \\
    & + \frac{1}{2 N^{2}} \frac{\partial^2}{\partial x^{2}}
    \left[\frac{\gamma}{N}\left(N n (2 - 4 \mu+s-\mu s)
    -n^{2}\left(2-4 \mu+s-2\mu s\right)\right) P(x, t)\right].
\end{split}
\end{equation}
Substituting the definition of the allele frequency results in
\begin{equation}
\begin{split}
    \frac{\partial}{\partial t} P(x, t)=
    &-\gamma \frac{\partial}{\partial x}[x s(1-x-\mu)+\mu(1-2 x) P(x, t)] \\
    &+\frac{\gamma}{2 N} \frac{\partial^{2}}{\partial x}
    \left[x(2-4 \mu+s-\mu s)-x^{2}(2-4 \mu+s-2 \mu s) P(x, t)\right].
\end{split}
\end{equation}
To get to the final equation we simply make use of the approximation that both
$s, \mu \ll 1$. Implementing this, and writing the time scale in units of
$\gamma^{-1}$ results in the classic diffusion theory equation with all three
forces implemented
\begin{equation}
    \frac{\partial}{\partial t} P(x, t) =
    -\frac{\partial}{\partial x}[s x(1-x) + \mu (1 - 2x) P(x, t)] 
    +\frac{1}{N} \frac{\partial^{2}}{\partial x^{2}}[x(1-x) P(x, t)].
\end{equation}